Ion pairing controls rheological properties of ``processionary'' polyelectrolyte hydrogels

### 2016-05-02 ion_pairing_paper.tex
Kuhn length Stretching of the eletrostatic blobs to obtain bounding energy from strain at crossing

As we will discuss in the following, all rheological features of the hydrogels studied here can  be rationalised on the basis of this ``processionary'' model. We base our analysis on two main observables, see Table~\ref{tab:results}: (i) the value of the shear modulus at small amplitude $G^\prime(\gamma\rightarrow 0)$, i.e. the elasticity of the undamaged gel network, and (ii) the critical strain $\gamma_c$ corresponding to the maximum of $G"$, that we consider as representative of the end of the linear regime and of the first broken cross-links\cite{Hyun2011}. By using these two parameters, we show that it is possible, with reasonable assumptions, to estimate microscopic parameters as the average number of chains between cross-links $n$, the head-to-body bonding energy $E_c$ or the number of monomers between dissociated ion pairs $A$.

@@

Let us first recall the structure of a polyelectrolyte chain as described by the scaling theory of \citet{Rubinstein1996}. In absence of strain, a linear polyelectrolyte chain is organised at various scales.

Below the thermal correlation length $\xi_T$ monomer interactions dominate over the thermal energy, thus chains behave ideally and we need to take into account the geometry of the polymer. Due to the relatively large lateral moieties ($s=11$ non-hydrogen atoms) with respect to the distance between them (2 carbons), the polymer can be considered as a comb polymer or a bottlebrush. According to \citet{Feuz2005} the persistence length is thus longer than the distance $b$ between two monomers and corresponds to the thickness $d=s^{2/3} a^{4/3}b^{-1/3}$ of the brush in ideal conditions, where $a\approx\SI{150}{\pico\metre}$ is the distance between two non-hydrogen atoms. Taking $b=a\sqrt{3}$, we obtain $d = \left(s/3\right)^{2/3}b \approx 2.4 b$. We note $g_c=d/b$ the number of monomers in the persistence length.

The density in thermal blobs is set by the reduced temperature $\tau= 1-T/\Theta$ such that $(g_T/g_c)/\xi_T^3 = \tau/d^3$ and their structure is a persistent random walk $\xi_T = d (g_T/g_c)^{1/2}$. These conditions yields $\xi_T = d/\tau$ and $g_T = g_c/\tau$.

At a scale $D$, the electrostatic energy balances the surface energy of the collapsed polymer. We name $g_e$ the number of monomers in the electrostatic blob. If there is a charge every $A$ monomers and there are $g_e/A$ charges in the electrostatic blob. The surface tension is proportional to the number of thermal blobs exposed on the surface $k_\mathrm{B}T/\xi_T^2$, with $k_\mathrm{B}$ is the Boltzmann constant and $T$ the temperature. Thus the balance of energies writes:
\begin{equation}
\left(\frac{D}{\xi_T}\right)^2 = \left(\frac{g_e}{A}\right)^2 \frac{\ell_\mathrm{B}}{D},
\label{eq:electrosurface}
\end{equation}
where $\ell_\mathrm{B} = e^2/(4\pi\epsilon k_\mathrm{B}T)$ is the Bjerrum length with $e$ the elementary charge, $\epsilon$ the dielectric constant of the solvent. In water $\ell_\mathrm{B} \approx \SI{0.7}{\nano\metre}$. Since thermal blobs fill the volume of the electrostatic blob $D = \xi_T \left(g_e/g_T\right)^{1/3}$, we obtain
\begin{equation}
g_e = \frac{A^2}{u} \tau, \qquad D = b\left(\frac{A^2}{u}g_c^2\right)^{1/3},\qquad \text{with }u = \ell_\mathrm{B}/b.
\end{equation}

We define the extension parameter $B$ as the ratio between the length of a chain of $g_e$ monomers fully extended and the actual size of the electrostatic blob: 
\begin{equation}
B = \frac{g_e b}{D} = \left(\frac{A^2}{u g_c}\right)^{2/3} \tau
\end{equation}

@@

Electrostatic repulsion decays with a characteristic length $r_\mathrm{src}$, the screening length. Following\cite{Rubinstein1996} this screening length is set by the concentration of dissociated counterions and the concentration of $c_s$ added salts:
\begin{equation}
r_\mathrm{src} = \left(\frac{B}{cb}\right)^{1/2} \left(1 + Bu \frac{c_s}{c}\right)^{-1/2}.
\end{equation}
Between $D$ and $r_\mathrm{src}$ the electrostatic blobs are organised in a linear rod containing $g_\mathrm{src}$ monomers. The length of this rod is $B$ times shorter than if the electrostatic blobs were extended: 
\begin{equation}
g_\mathrm{src} = B \frac{r_\mathrm{src}}{b} = \left(\frac{B^3}{cb^3}\right)^{1/2} \left(1 + Bu \frac{c_s}{c}\right)^{-1/2}.
\label{eq:gscr}
\end{equation}

@@

\begin{table}
\begin{tabular}{l|SS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_0$} & {$n$} & {$B$} & \multicolumn{2}{c}{$A$}\\
&	{\si{\pascal}} & {\%} & & &{$\tau=0.2$}& {$\tau=1$}\\\hline&&&&&\\[-10pt]
\ce{PIm+Br-}	& 13	&	1.1	&	860	&	740	&	1370	&	410\\
\ce{PIm+I-}		& 78	&	1.8	&	122	&	104.8	&	317	&	94\\
\ce{PPyr+Cl-}	& 270 	&	2.8	&	50.6	&	55.7	&	197	&	59\\
\ce{PPyr+Br-}	& 9800 	& 	0.3	&	1.15	&	2.91	&	21	&	6.5\\
\ce{PPyr+I-}	& 8800	&	0.0	&	1.08	&	2.63	&	19	&	6.0\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates inconsistent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

For all our samples, the linear regime is narrow $\gamma_0\approx 10^{-2}\ll 1$ if existing. It means that procession is just long enough to start performing the self-avoiding walk. Incidentally, this implies that the procession is too short to perform the random walk. In other words the number $N$ of monomer between crosslinks is comparable to $g_\mathrm{scr}$. Using equation (\ref{eq:gscr}), this condition translates into
\begin{equation}
\left(\frac{B}{B_0}\right)^3 = 1 + \frac{B}{B_s}
\label{eq:solveB}
\end{equation}
with $B_s = c/u c_s$ and $B_0 = b c^{1/3} N^{2/3} = bc \left(k_\mathrm{B}T/G\right)^{2/3}$. Here, the ``salt'' are the phosphonate heads, that is to say two charges per chain. Thus $c_s/c = 2/N_0$ and $B_s = N_0/2u \approx 23$.

@@

Numerical solutions of Equation (\ref{eq:solveB}) are given in Table~\ref{tab:results}. For the same quality of solvent $B$ is monotonically related to the amount of counterion condensation. We thus confirm that our samples are sorted by decreasing counterion condensation (decreasing $B$). More quantitatively, the number $A$ of monomers between dissociated charges is
\begin{equation}
A = g_c u^{1/2} \left(\frac{B}{\tau}\right)^{3/4}
\end{equation}
We know that $\Theta>\SI{100}{\celsius}$ so $0.2<\tau<1$.

\subsubsection{Bounding energy}

Considering that between two cross-links a procession is an entropic spring in its linear domain, the mechanical energy stored under a strain $\gamma$ is $E = \gamma^2 k_\mathrm{B}T$. If we suppose that ionic bonds break before the complete extension of the procession, we evaluate the head-to-body bonding energy to
\begin{equation}
E_c = \gamma_c^2 k_\mathrm{B}T.
\label{eq:Ec}
\end{equation}

For \ce{PIm+Br-} we measure $\gamma_c = 800\%$ and thus $E_c \approx 65 k_\mathrm{B}T$ per head-to-body bond or \SI{160}{\kilo\joule\per\mol}. This value is much higher than the $E_i\approx\SI{5}{\kilo\joule\per\mol}$ typical of ionic bonds in water\cite{Schneider1992}, even when adding up the two bonds of the bivalent phosphonate head\bibnote{Since the two minus charges are borne by the same atom, chelation effect may enhance bounding energy, but not by an order of magnitude}. However, one has to keep in mind that the chain is in bad solvent, thus forming locally compact blobs of low water content. Therefore, the medium surrounding the ionic bond is not water but the hydrophobic chains. Such a low polarity microenvironement is well known to enhance otherwise weak electrostatic interactions in protein folding or engineered self-assembly\cite{Rehm2010}. Quantitatively, the bonding energy of an ionic bond is inversely proportional to the relative dielectric constant $\epsilon_r$, therefore we need to suppose a local $\epsilon_r$ around 5 (between 2-3 typical of water-insoluble polymers and 80 for water) to recover the large bonding energy that we measure.

However for all other samples the critical strain is between 2\% and 20\%, leading to inconsistent values for the bonding energies ranging between $10^{-3}k_\mathrm{B}T$ and $10^{-1}k_\mathrm{B}T$ per bond, see Table~\ref{tab:results}. Therefore, the end of the linear regime cannot correspond to bond breaking but rather to the point where extended chains become nonlinear.

\subsubsection{From critical strain to charge dissociation}

The spatial conformation of a polyelectrolyte chain in semi-dilute, low salt solution is a persistent random walk where the persistence length is the screening length $r_\mathrm{src}$ due to the dissociated counter ions\cite{Rubinstein1996}. In bad solvent, the persistent random walk is made of electrostatic blobs of diameter $D$ containing $g_e$ monomers, see Figure~\ref{fig:dissociation}a:
\begin{align}
g_e &= \frac{A^2}{u}\tau\label{eq:ge},\\
D &= b \left(\frac{A^2}{u}\right)^{1/3}\label{eq:D}.
\end{align}
Here $A$ is the number of monomers between dissociated ion pairs, $\tau = 1 - T/\Theta$ the reduced temperature and $u = \ell_\mathrm{B}/b \approx 2.7$ is the ratio of the Bjerrum length in water $\ell_\mathrm{B}\approx \SI{0.7}{\nano\metre}$ and monomer size $b\approx \SI{0.26}{\nano\metre}$.

Extending these electrostatic blobs means exposing more monomers to the bad solvent and typically requires energies at least an order of magnitude higher than for the extension of the persistent random walk. We will thus consider that the exit from the linear domain of the material corresponds to the full extension of the persistent random walk:
\begin{equation}
\gamma_c = \frac{R_0}{R} - 1
\end{equation}
where $R$ and $R_0$ are respectively the end-to-end distance and the contour length of the persistent random walk.

A segment of the persistent random walk is $r_\mathrm{scr}$ long and thus contains $g_\mathrm{scr} = g_e r_\mathrm{scr}/D$ monomers. We thus have
\begin{equation}
\left. \begin{array}{ll}
R &= r_\mathrm{scr} \left(\frac{N}{g_\mathrm{scr}}\right)^{1/2}\\
R_0 &= r_\mathrm{scr} \frac{N}{g_\mathrm{scr}}
\end{array}\right\rbrace\quad
\gamma_c = \left(\frac{N}{g_\mathrm{scr}}\right)^{1/2} -1\label{eq:gamma0}.
\end{equation}

\citet{Rubinstein1996} express the screening length as
$r_\mathrm{scr} = \left(A^2 / u\right)^{1/3} \left(\tau /cb \right)^{1/2}$. Combining with (Eq.~\ref{eq:ge}) and (Eq.~\ref{eq:D}) we deduce
\begin{equation}
g_\mathrm{scr} = \frac{A^2}{u} \left(\frac{\tau^3}{c b^3}\right)^{1/2}\label{eq:gscr}.
\end{equation}

From (Eq.~\ref{eq:gamma0}) and (Eq.~\ref{eq:gscr}) we can finally relate the measured critical strain to the microscopic parameter $A$:
\begin{equation}
A = \frac{\left(N u\right)^{1/2}}{\gamma_c+1}\left(\frac{c b^3}{\tau^3}\right)^{1/4}.
\label{eq:A}
\end{equation}

The only unknown here is the reduced temperature $\tau$, possibly different between poly\-(pyrrolidinium) based and poly\-(immidazolium) based polymers. We know that the undissociated chain has a $\Theta$ temperature higher than \SI{100}{\celsius}, thus $0.2<\tau<1$. In addition, we must have $A\geq 1$, which implies that $\tau\leq 0.235$ through (Eq.~\ref{eq:A}), corresponding to $\Theta\approx\SI{116}{\celsius}$. This limit value yields $A=1$ for both \ce{PPyr+I-} and \ce{PPyr+Br-}, indicating full dissociation of the cationic moieties. By contrast \ce{PPyr+Cl-} has a charge every 7 monomers. Supposing the same $\tau$, \ce{PIm+I-} has a dissociated ion pair every 10 monomers, see Table~\ref{tab:results}. Of course analysing \ce{PIm+Br-} based on the value of $\gamma_c$ gives an inconsistent value $A=3.4$ indicating that this hydrogel does not belong to the same regime.

\subsubsection{Charge dissociation and cross-linking ratio}

Finally we cross-check our model by linking the ion pair dissociation and the amount of physical cross-links by an independent argument. A second charge dissociated on a chain creates a possible binding site for a second head-to-body association and thus a cross-link. Therefore the probability to have a cross link is proportional to the probability of having a second charge dissociation. If we assume that the probability for an ion pair to dissociate is independent of the number of dissociated pairs already present on the polymer, we can describe the probability of having $k$ dissociated pairs on a polymer by a Poisson distribution of mean $\lambda = N_0/A$. Therefore, the cross-link probability $1/n$ is proportional to $\lambda^2$, yielding $n \sim A^2$.

As shown in Figure~\ref{fig:dissociation}b and the last two columns of Table~\ref{tab:results} this relation describes very well our data with a prefactor close to 1. We can extrapolate this law for the case of \ce{PIm+Br-} and obtain $A = 28$. Inverting (Eq.~\ref{eq:A}) yields $R_0/R - 1 = 860\%$, indeed higher than the observed $\gamma_c\approx 804\%$. This confirms that in the \ce{PIm+Br-} strain sweep $\gamma_c$ actually results from the strength of the ionic bonds.

%We can now estimate the hydrophobic interaction energy $E_h$ by considering that to separate two chains two interfaces with water of area $\pi D^2/4$ have to be created. Surface tension is $\tau^2 k_\mathrm{B}T / b^2$\cite{Rubinstein1996}. Combining with (Eq.~\ref{eq:D}) we find
%\begin{equation}
%\frac{E_i}{k_\mathrm{B}T} = \frac{\pi}{2}\tau^2\left(\frac{A^2}{u}\right)^{2/3} \approx 4
%\end{equation}

%\subsection{Discussion}
%
%The above dependence of the gel parameters on the nature of the counterions could be rationalized on the basis of ion-pairing processes. The unsolubility/unswellability of \ce{PIm+F-}, \ce{PPyr+F-} and \ce{PIm+Cl-} suggest that ion pairs are not efficiently dissociated in aqueous medium, thus hindering the solvation process. By comparison, \ce{PIm+Br-}, \ce{PIm+I-}, \ce{PPyr+Br-} and \ce{PPyr+I-} form homogeneous materials when suspended in water, which one can correlate with efficient dissociation of the halide counterions from the polymer main chain is fully consistency with decreased ionic bond strength. Such dissociation, in turns makes it possible for the ``hard'' phosphonate group on each polymer chain's end to interact with the ``free'' cationic groups of the side chains, which we showed in a previous paper to be at the origin of the gel formation mechanism\cite{Srour2014}.
%
%The similarity in gel parameters observed for \ce{PPyr+Br-} and \ce{PPyr+I-} ($G^\prime\approx\SI{8500}{\pascal}$) suggests that full ion pair dissociation is achieved in the latter, or at least that this dissociation is enough in both cases to provide maximal interaction between the terminal phosphonate and the pyrrolidinium side groups. Conversely the significantly different behavior between \ce{PIm+I-} ($G^\prime= \SI{78}{\pascal}$) and \ce{PIm+Br-} ($G^\prime= \SI{13}{\pascal}$) indicates that dissociation is only partial and occurs to a lesser extent in the latter than in the former. The same conclusion can be drawn for \ce{PPyr+Cl-}, in which the elastic modulus ($G^\prime= \SI{270}{\pascal}$) value drops by more than one order of magnitude as compared to its bromide and iodide counterparts.
%
%As already mentioned (vide supra) another striking feature of these gels is that in similar conditions poly(pyrrolidinium) based gels are systematically much stronger than those formed with poly(imidazolium). Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge; consequently, according to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected for the latter with the ``soft'' halide counterions\cite{Goossens2009}. As a consequence, for a given halide, dissociation is less favored in the case of poly(imidazolium), as clearly illustrated by the difference in solvation between \ce{PIm+Cl-} and \ce{PPyr+Cl-}. Conversely, association of the lateral cations with the ``hard'' phosphonate moieties at the polymer chain’s end is expected to be stronger with pyrrolidinium, hence a systematically larger elastic modulus. The interplay of these two effects contributes in making poly(pyrrolidinium) based gels much stronger than their poly(imidazolium) counterparts; comparison between \ce{PIm+Br-} and \ce{PPyr+Br-} is particularly illustrative of this trend, as their $G^\prime$ value (13 and \SI{8100}{\pascal}, resp.) differ by almost three orders of magnitude.

>>>>

As we will discuss in the following, all rheological features of the hydrogels studied here can  be rationalised on the basis of this ``processionary'' model. We base our analysis on three main observables, see Table~\ref{tab:results}: (i) the value of the shear modulus at small amplitude $G^\prime(\gamma\rightarrow 0)$, i.e. the elasticity of the undamaged gel network, (ii) the strain amplitude $\gamma_0$ corresponding to the end of the linear regime~\cite{Hyun2011}, and (iii) the strain amplitude $\gamma_c$ corresponding to the crossing of the moduli. By using these three parameters, we show that it is possible, with reasonable assumptions, to estimate microscopic parameters as the average number of chains between cross-links $n$, the number of monomers between dissociated ion pairs $A$ and the head-to-body bonding energy $E_c$.

@@

The scaling theory of \citet{Rubinstein1996} describes the structure of a polyelectrolyte chain. In the following, we make the link between this theory and the chemical structure of our polymers. In absence of strain, the structure of a polyelectrolyte chain is organised at various scales.

The smallest scale is the Kuhn length $b$ containing $g_\mathrm{K}$ monomers. For a freely joint chain of tetrahedral carbons $b\approx 367$ pm which corresponds to 3 tetrahedral carbons. Since a monomer counts 2 carbons along the chain $g_\mathrm{K}\approx 3/2$.

In bad solvent, the thermal length $\xi_T$ is larger than $b$. The thermal blob contains $g_T$ monomers and its density is set by the reduced temperature $(g_T/g_\mathrm{K})/\xi_T^3 = \tau/b^3$. Between $b$ and $\xi_T$ we have a persistent random walk $\xi_T = b (g_T/g_\mathrm{K})^{1/2}$. These conditions yields $\xi_T = b/\tau$ and $g_T = g_\mathrm{K}/\tau^2$.

At a scale $D$, the electrostatic energy balances the surface energy of the collapsed polymer. We name $g_e$ the number of monomers in the electrostatic blob. If there is a charge every $A$ monomers then there are $g_e/A$ charges in the electrostatic blob. The surface tension is proportional to the number of thermal blobs exposed on the surface $k_\mathrm{B}T/\xi_T^2$, with $k_\mathrm{B}$ the Boltzmann constant and $T$ the temperature. Thus the balance of energies writes:
\begin{equation}\left(\frac{D}{\xi_T}\right)^2 = \left(\frac{g_e}{A}\right)^2 \frac{\ell_\mathrm{B}}{D},
\label{eq:electrosurface}
\end{equation}
where $\ell_\mathrm{B} = e^2/(4\pi\epsilon k_\mathrm{B}T)$ is the Bjerrum length with $e$ the elementary charge, $\epsilon$ the dielectric constant of the solvent. In water $\ell_\mathrm{B} \approx \SI{0.7}{\nano\metre}$. Since thermal blobs fill the volume of the electrostatic blob $D = \xi_T \left(g_e/g_T\right)^{1/3}$, we obtain
\begin{equation}
g_e = \frac{A^2}{u g_\mathrm{K}} \tau, \qquad D = b\left(\frac{A^2}{u g_\mathrm{K}^2}\right)^{1/3},\qquad \text{with }u = \ell_\mathrm{B}/b.
\label{eq:geD}
\end{equation}

We define the extension parameter $B$ as the ratio between the length of a chain of $g_e$ monomers fully extended and the actual size of the electrostatic blob: 
\begin{equation}
B = \frac{g_e}{g_\mathrm{K}}\frac{b}{D} = \left(\frac{A^2}{u}\right)^{2/3} g_\mathrm{K}^{-1/3}\tau
\label{eq:B}
\end{equation}

@@

In dilute solution, the chain is a linear rod of length $L= N b / B g_\mathrm{K}$. The overlap concentration is thus $c^* = N/L^3 = g_\mathrm{K}B / b L^2$. Following \cite{Dobrynin1995}, for a monomer concentration $c > c^*$ and in absence of salt the screening length is
\begin{equation}
r_\mathrm{src} = L \left(\frac{c^*}{c}\right)^{1/2} =  \left(\frac{B g_\mathrm{K}}{cb}\right)^{1/2}.
\label{eq:rscrNoSalt}
\end{equation}

In the presence of added salt, electrostatic interactions are screened at the Debye length $r_\mathrm{src} = \left(\ell_\mathrm{B} c_s\right)^{-1/2}$. For arbitrary salt concentration,
\begin{equation}
r_\mathrm{src} = \left(\frac{B g_\mathrm{K}}{cb}\right)^{1/2} \left(1+ B u \frac{c_s}{c} g_\mathrm{K}\right)^{-1/2}.
\label{eq:rscr}
\end{equation}

Between $D$ and $r_\mathrm{src}$ the electrostatic blobs are organised in a linear rod containing $g_\mathrm{src}$ monomers. The length of this rod is $B$ times shorter than if the electrostatic blobs were extended:
\begin{equation}
g_\mathrm{src} = B \frac{r_\mathrm{src}}{b} g_\mathrm{K} = \left(\frac{B^3 g_\mathrm{K}^3}{cb^3}\right)^{1/2} \left(1 + B u \frac{c_s}{c} g_\mathrm{K}\right)^{-1/2}.
\label{eq:gscr}
\end{equation}

@@

\begin{table}
\sisetup{
separate-uncertainty,
%table-format = 1.3(1),
}
\begin{tabular}{l|SSS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_0$} & {$\gamma_c$} & {$n$} & {$B$} & {$A$} & {$E_c$} & {$E_c$}\\
&	{\si{\pascal}} & {\%} & {\%} & & & & {\si{\kilo\joule\per\mole}}& {$k_\mathrm{B}T$}\\\hline&&&&&\\[-10pt]
\ce{PIm+Br-}	& 13	&	1.1	&	810	&	860	&	930	&	383	&	300	&	120\\
\ce{PIm+I-}		& 78	&	1.8	&	18	&	122	&	126	&	85	&	116	&	47\\
\ce{PPyr+Cl-}	& 270 	&	2.8	&	12	&	50.6	&	64	&	52	&	67	&	27\\
\ce{PPyr+Br-}	& 9800 	& 	0.3	&	2	&	1.15	&	2.8	&	5	&	7.3	&	2.9\\
\ce{PPyr+I-}	& 8800	&	0.0	&	89	&	1.08	&	2.5	&	4.6	&	7.2	&	2.9\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. $A$ and $E_c$ are obtained supposing $\tau=0.588$.}
\label{tab:results}
\end{table}

@@

For all our samples, the linear regime is narrow $\gamma_0\approx 10^{-2}\ll 1$ if existing. It means that procession is just long enough to start performing the self-avoiding walk. Incidentally, this implies that the procession is too short to perform the random walk. In other words the number $N$ of monomers between crosslinks is comparable to $g_\mathrm{scr}$. Using equation (\ref{eq:gscr}), this condition translates into
\begin{equation}
\left(\frac{B}{B_0}\right)^3 = 1 + \frac{B}{B_s}
\label{eq:solveB}
\end{equation}
with $B_s = c/u c_s g_\mathrm{K}$ and $B_0 = b c^{1/3} N^{2/3} g_\mathrm{K}^{-1} = (bc/g_\mathrm{K}) \left(k_\mathrm{B}T/G\right)^{2/3}$. Here, the ``salt'' consists in the phosphonate heads, that is to say two charges per chain. Thus $c_s/c = 2/N_0$ and $B_s = N_0/2u g_\mathrm{K} \approx 12$.

@@

Numerical solutions of Equation (\ref{eq:solveB}) are given in Table~\ref{tab:results}. For the same quality of solvent $B$ is monotonically related to the amount of counterion condensation. We thus confirm that our samples are sorted by decreasing counterion condensation (decreasing $B$). More quantitatively, the number $A$ of monomers between dissociated charges is
\begin{equation}
A = u^{1/2} \left(\frac{B}{\tau}\right)^{3/4} g_\mathrm{K}^{1/4}
\label{eq:A}
\end{equation}.

For the theory to be correct, we need the electrostatic length to be larger than the thermal length, in other words $g_e \geq g_T$. Using Equations (\ref{eq:geD}) and (\ref{eq:B}), we express $g_e$ function of $B$ and $\tau$:
\begin{equation}
g_e = \left(\frac{B^3}{g_\mathrm{K}\tau}\right)^{1/2},
\end{equation}
which yields the condition $\tau \geq \frac{g_\mathrm{K}}{B}$. This condition it the most stringent for PPyrI (smallest $B$) and yields $\tau \geq 0.58$. In the following we set $\tau$ to this minimal value that corresponds to $\Theta=\SI{450}{\celsius}$.

The values we compute for $A$ using Equation (\ref{eq:A}) are shown in Table~\ref{tab:results}. $A$ varies between $4.6$ for \ce{PPyr+I-}, corresponding to $15$ uncondensed counterions per polymer, to $383$ for \ce{PIm+Br-} indicating less than a single dissociated counterion every 5 polymers.

\subsubsection{Extention of the electrostatic blobs}

The fully extended self-avoiding walk is a cylinder of length $R_0 = Nb/Bg_\mathrm{K}$, of diameter $D$, of volume $V = R_0 D^2$ and of area $\Sigma_0 = R_0 D = (R_0 V)^{1/2}$. Extending the cylinder to $R = (\gamma+1) R_0$ while keeping the volume $V$ constant thus creates an extra area 
$
\Delta\Sigma = \left(\left(\gamma +1\right)^{1/2} -1\right) \Sigma_0
$
that translates into a stretching energy due to the surface tension $k_\mathrm{B}T/\xi_T^2$:
\begin{equation}
\frac{E_\mathrm{stretch}}{k_\mathrm{B}T} = \left(\left(\gamma +1\right)^{1/2} -1\right) N \left(\frac{\tau^3}{B g_\mathrm{K}^5}\right)^{1/2}.
\label{eq:stretch}
\end{equation}

If we consider that the head-to-body bonds break at $\gamma_c$, we can extract their bonding energy $E_c$ from Equation (\ref{eq:stretch}). The results are reported in Table~\ref{tab:results}. In the most dissociated \ce{PPyr+I-} and \ce{PPyr+Br-}, the head-to-body bonding energy $E_c$ is compatible with typical ionic bonds, \SI{5}{\kilo\joule\per\mole} in water~\cite{Schneider1992}. However the bonding energy increases with increasing counterion condensation to reach 85\% of the carbon-carbon single bond (\SI{350}{\kilo\joule\per\mole} or $140 k_\mathrm{B}T$) in the case of \ce{PIm+Br-}. Indeed, at high counterion condensation, the electrostatic blobs are large, up to 530 polymers for \ce{PIm+Br-}, and embed head-to-body bonds into compact blobs of low water content. Such a low polarity microenvironement is well known to enhance otherwise weak electrostatic interactions in protein folding or engineered self-assembly~\cite{Rehm2010}.
